raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-10-13 code/rmarkdown/article.Rmd
+ assorted fixes

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r9]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@r476]. Additionally, migratory processes allow many species to access suitable habitat throughout the year [r477]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @r406; @r33; @r460; @r453]. This is sparked significant interest in using this data to make better conservation decisions [@r473; @r9; @r448; @r2; @r68; @r470; @r469; @r247; @r488; @r468; @r407; @r61; @r475].

<!-- current methods for preserving biodiversity processes -->
Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Marxan} and \texttt{Zonation}; @r22; @r429]. The cornerstone of these tools is that decision makers can set targets for a proportion of the geographic range of species, or ecosystems (termed features). To incorporate some information on biodiversity processes, current methods involve partitioning the features into different groups as a pre-processing step [@r9; @r448; @r247]. These groups are partitioned based on a proxy variable that relates to a biodiversity process of interest [@r473]. For instance, species ranges' have been partitioned according to habitat discontinuities to capture adaptive processes [@r2]. However, by partitioning features into different groups, it is not clear how well a given solution samples the original proxy variable, and ultimately, how well the solution captures the biodiversity process of interest. To overcome this limitation, Faith and Walker [-@r218] developed a decision support tool (\texttt{DIVERSITY}) to explicitly sample proxy variables to secure biodiversity processes. However, this tool is not widely used today, in part, because it cannot accommodate explicit targets on biodiversity patterns for multiple features.

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of unified a decision support tool that uses explicit and separate targets for biodiversity patterns and processes. To begin to fill this gap, we united the ideas underpinning \texttt{DIVERSITY} and \texttt{MARXAN} into new optimisation problems, and implemented them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of its inner workings and explore its functionality.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

@@

Demand points are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space. The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

@@

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Supporting Information S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.

>>>>

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r9]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@r476]. Additionally, migratory processes allow many species to access suitable habitat throughout the year [@r477]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @r406; @r33; @r460; @r453]. This is sparked significant interest in using this data to make better conservation decisions [@r68; @r9; @r473; @r247; @r448; @r468; @r407; @r61; @r2; @r470; @r475; @r469].

<!-- current methods for preserving biodiversity processes -->
Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Marxan} and \texttt{Zonation}; @r429; @r22]. The cornerstone of these tools is that decision makers can set targets for a proportion of the geographic range of species, or ecosystems (termed features). To incorporate some information on biodiversity processes, current methods involve partitioning the features into different groups as a pre-processing step [@r9; @r247; @r448]. These groups are partitioned based on a proxy variable that relates to a biodiversity process of interest [@r473]. For instance, species ranges' have been partitioned according to habitat discontinuities to capture adaptive processes [@r2]. However, by partitioning features into different groups, it is not clear how well a given solution samples the original proxy variable, and ultimately, how well the solution captures the biodiversity process of interest. To overcome this limitation, Faith and Walker [-@r218] developed a decision support tool (\texttt{DIVERSITY}) to explicitly sample proxy variables to secure biodiversity processes. However, this tool is not widely used today, in part, because it cannot accommodate explicit targets on biodiversity patterns for multiple features.

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unified decision support tool that uses explicit and separate targets for biodiversity patterns and processes. To begin to fill this gap, we united the ideas underpinning \texttt{DIVERSITY} and \texttt{MARXAN} into new formulations of the reserve selection problem, and implemented them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore the functionality.

# Methods
## OPTIMISATION PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

@@

Demand points are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space (see below for mathematical expressions). The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

@@

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are named after the unreliable and reliable facility location formulations from which they are based upon [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

@@

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (eqn 3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{raptr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations (see Appendix S1 for formulation). These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website.
